---
title: "Day Wrangling in Tidyverse tidyr"
author: "Erica Dale"
date: "2022-08-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(janitor)
library(here)
```

## Read In World Bank data files

```{r}
wb_indicators <- read_csv(here("data", "wb_indicators.csv"), na = c("..", ""))

# View the data
### Unknown values are .. and "" which are auto-read as characters for a column

wb_metadata <- read_csv(here("data", "wb_indicators_metadata.csv"))
```

## Tidy up data

"Years" column titles: needs own column "Series Name" column: separate into different variables

### Combine the years columns

```{r}
wb_indicators_long <- pivot_longer(wb_indicators, 
                                   cols = '2001 [YR2001]':'2020 [YR2020]', 
                                   names_to = "year",
                                   values_to = "indicator_value"
                                   )
names(wb_indicators)
# OR
wb_indicators_long <- wb_indicators %>% pivot_longer(cols = -(1:4),
                               names_to = "year",
                               values_to = "indicator_value")
```

### Separate to clean up the year column

```{r}
wb_separate <- wb_indicators_long |> 
  separate(col = year, into = c("year", "year_chr"), sep = " ") |> 
  select(-year_chr, -'Series Code', - 'Country Code')

class(wb_separate$year) # Still character value
glimpse(wb_separate)

# Update year class

wb_separate <- wb_indicators_long |> 
  separate(col = year, into = c("year", "year_chr"), sep = " ") |> 
  select(-year_chr, -'Series Code', - 'Country Code') |> 
  mutate(year = as.numeric(year))
```

### Separate Series column

```{r}
unique('Series Name')

wb_clean <- wb_separate |> 
  drop_na('Series Name') |>   # NEED TO DROP NA for column names
  pivot_wider(names_from = 'Series Name', values_from = indicator_value)

```

### Rename column names

Rename from dplyr (great if just one) Janitor if simple column names

```{r}
names(wb_clean)
names(wb_clean) <- c("country", "year", "access_clean_fuels_pp", "access_electricity_pp", "co2_emissions_kt", "fossil_fuel_cons_pct", "water_stress")

# Names cannot be piped into

wb_clean_2 <- wb_clean |> 
  rename(the_year = year,
         the_country = country)
```

### Example of filtering

This is why tidy data is useful

```{r}
wb_subset <- wb_clean |> 
  filter(country %in% c("Sweden", "Thailand", "Trinidad and Tobago", "Botswana"))

ggplot(data = wb_subset, aes(x = year, y = co2_emissions_kt)) +
  geom_line(aes(color = country)) +
  facet_wrap(~country)
```

## Filtering Rows

### Filter to only keep exact matches

```{r}
ex_1 <- starwars |> 
  filter(height > 180)
min(ex_1$height)
```

```{r}
ex_2 <- starwars |> 
  filter(eye_color == "yellow")
unique(ex_2$eye_color)
```

```{r}
ex_3 <- starwars |> 
  filter(homeworld == "Naboo")
unique(ex_3$homeworld)
```

### Filter to satisfy multiple conditions

AND statements
```{r}
ex_4 <- starwars |> 
  filter(height > 180 & homeworld == "Tatooine")
unique(ex_4$homeworld)
unique(ex_4$height)
```

```{r}
ex_5 <- starwars |> 
  filter(hair_color == "brown" & species == "Human")
unique(ex_5$hair_color)
unique(ex_5$species)
```

OR statements
```{r}
ex_6 <- starwars |> 
  filter(height > 180 | eye_color == "yellow")
unique(ex_6$eye_color)
unique(ex_6$height)
```

```{r}
ex_7 <- starwars |> 
  filter(homeworld == "Endor" | species == "Droid")
```

Writing a different way when referring to multiple character values
```{r}
ex_8 <- starwars |> filter(eye_color %in% c("blue", "brown", "red"))
unique(ex_8$eye_color)

# DO NOT USE == , will search IN ORDER of one vector value with each row instead of asking for all of them
```

```{r}
ex_11 <- starwars |> 
  filter(homeworld %in% c("Endor", "Tatooine", "Naboo"),
         height < 150)
view(ex_11)
```

#### Negate Operator
```{r}
ex_12 <- starwars |> 
  filter(species != "Human")
unique(ex_12$species)
```

## Select or exclude columns

Select by name
```{r}
ex_14 <- wb_clean |> 
  select(country, year, co2_emissions_kt)
names(ex_14)
```

Select range of names
```{r}
ex_15 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct)
# Risky do to because range is positional and columns can be moved around
names(ex_15)
```

Exclude one from range
```{r}
ex_16 <- wb_clean |> 
  select(year:fossil_fuel_cons_pct, - access_electricity_pp)
names(ex_16)
```

Use it to rename/move things around
```{r}
ex_18 <- wb_clean |> 
  select(year_new = year,
         emissions = co2_emissions_kt,
         country_name = country)
glimpse(ex_18)
```

### Pipe some together
```{r}
ex_17 <- wb_clean |> 
  filter(country %in% c("Bolivia", "Chile", "Mexico")) |> 
  select(country:fossil_fuel_cons_pct)
names(ex_17)
unique(ex_17$country)
```

Practice
```{r}
ex_19 <- wb_clean |> 
  select(year, country, water_stress) |> 
  filter(country %in% c("Cambodia", "South Africa", "Nepal"))
unique(ex_19$country)
glimpse(ex_19)
```


### Rename columns

Select requires many selections for renaming
```{r}
ex_20 <- wb_clean |> 
  rename(ffcp = fossil_fuel_cons_pct,
         ws = water_stress)
ex_21 <- wb_clean |> 
  rename(co2_e_kt = co2_emissions_kt,
         the_country = country)
```

### Mutate columns
If you assign a NEW column name, it will add a new column to the end of the data frame.
If assign an EXISTING column name, it will overwrite the existing column.

```{r}
ex_22 <- wb_clean |> 
  mutate(co2_emissions_t = co2_emissions_kt * 1000) |> 
  relocate(co2_emissions_t, .after = co2_emissions_kt)

ex_23 <- wb_clean |> 
  mutate(yay = "YAY")

ex_24 <- wb_clean |> 
  mutate(year = as.character(year))
class(ex_24$year)



```

### Relocate columns
dplyr::relocate
Can move conditionally by classes (ie. all numeric after all character columns)
```{r}
ex_25 <- wb_clean |> 
  relocate(water_stress, .after = year)
```


## Groupby Summarize

Group_by: Creates groups under the hood based on variables specified
Summaraize: apply functions to each GROUP and return in summary table

```{r}
ex_27 <- starwars |> 
  filter(homeworld %in% c("Naboo", "Tatooine")) |> 
  group_by(species, homeworld) |> 
  summarize(mean_height = mean(height, na.rm = TRUE),
            mean_mass = mean(mass, na.rm = TRUE))
view(ex_27)
```

Make a data frame
Frequency table
Case table - each observation gets its own row
```{r}
df <- tribble(
  ~species, ~length, ~number,
  "lobster", 12, 4,
  "lobster", 14, 6,
  "lobster", 15, 2
)
df_case <- df |> 
  uncount(number)
df_og <- df_case |> 
  count(species, length)
# count is a combo of group_by, summarize, n

starwars |> 
  count(homeworld, species)
# THese two are the same
starwars |> 
  group_by(homeworld, species) |> 
  summarize(size = n())
```

```{r}
ex_30 <- wb_clean |> 
  filter(country %in% c("Nicaragua", "Ecuador", "Peru", "Chile"),
         year > 2005) |> 
  select(country, year, co2_emissions_kt) |> 
  mutate(co2_em_mg = co2_emissions_kt * 1000) |> 
  group_by(year) |> 
  summarize(sum(co2_em_mg))

```

